Fourier-domain modulations and delays of gravitational-wave signals

### 2018-02-14 docs/paper_method.tex
Significant edits of the Intro and moderate edits of Sec II.

With the recent detections of gravitational-wave signals emitted by the coalescences of binary black holes, announced by the LVC collaboration during the first and second obervation runs of the two LIGO instruments~\cite{LIGO-theevent-2016,LIGO-christmasevent-2016,LIGO-O1BBH-2016}\jgb{[update refs]}, gravitational-wave astronomy has entered its observational era. With the conclusion of LIGO's second observation run, and the recent extension of the ground-based detectors network with Virgo~\cite{Virgo} ( and later also KAGRA~\cite{KAGRA} and LIGO-India~\cite{INDIGO}), many more detections from compact object coalescences are to come, with binary black holes being expected to be the dominant sources~\cite{} \SM{[C]}.

Moreover, the European Space Agency has issued a mission call \SM{[update]} for its L3 space mission with a tentative launch around 2034, adressing the science theme ``The Gravitational Universe'', possibly opening the way for a space-borne observatory based on the concept of the Laser Interferometer Space Antenna (LISA)~\cite{LISA09,LISA17}. Among other sources of gravitational waves, LISA will be able to detect and characterize comparable-mass binary black hole coalescences to cosmological distances over a wide range of masses, from high-redshift observations of supermassive black hole binaries with $M\sim 10^{7} \Msol$ down to the observation of LIGO-type sources with $M\sim 10^{2} \Msol$~\cite{Sesana16}.

Data analysis for gravitational-wave observations of compact binary coalescences require accurate models (or templates) for the signals, both for ensuring efficient detections of signals that may be buried in instrumental noise, and to extract the physical parameters of the source in a subsequent analysis. The the greater sensitivity of LISA and other future instruments will require further increases in the accuracy of computationally efficient signal templates. This motivates an important and ongoing effort to improve gravitational wave templates. Two important focus areas for template improvements are need to cover all phases of the evolution of the system, including the inspiral, merger and ringdown (IMR), which became possible thanks to the breakthrough of numerical relativity~\cite{Pretorius05,Baker+06,Campanelli+06}, and the inclusion of the effects of the spins of the compact objects, that can lead to a precession of the orbital plane~\cite{ACST94,K95}. Besides accuracy, the computational efficiency of template generation is crucial. Bayesian analysis for parameter estimation of gravitational-wave signals, as was performed for the LIGO detections~\cite{LIGO-theeventPE-2016,LIGO-O1BBH-2016}, may require millions of evaluations of the likelihood function to sample the posterior probability distribution. \SM{[objective: accuracy, not necessary necessary for LIGO now but in the future yes]}

The GW community is making progress assembling higher-fidelity tools for these coming challenges.
State-of-the-art IMR templates combine information from the perturbative results of post-Newtonian theory covering the inspiral (see e.g.~\cite{BlanchetLiving}) and from numerical relativity simulations covering the end of the inspiral and the merger-ringdown phase (see e.g.~\cite{Pfeiffer12}). The Effective-One-Body (EOB) approach~\cite{BD99} does so by incorporating a resummation of post-Newtonian results and a calibration to numerical relativity, and has been extended to include both spins of the two bodies~\cite{Taracchini+13, Pan+13, Bohe+16}. Phenomenological templates extend post-Newtonian models by postulating an analytic ansatz for the Fourier-domain amplitude and phase, fitting coefficients against numerical relativity waveforms~\cite{Husa+15, Khan+15}, and can include a simplified model for precession~\cite{Hannam+13}. The second type of models produce directly the amplitude and the phase of the Fourier-domain waveform components in the form of a piecewise analytic function. EOB waveforms, by contrast, require integration of ordinary differential equations in the time-domain, which can be time-consuming. For the non-precessing case, Reduced Order Models (ROM), also called surrogate models, have been developed to speed up considerably the waveform generation, without losing accuracy~\cite{Field+13, Puerrer14, Bohe+16}. Both approaches are therefore able to efficiently produce IMR Fourier-domain waveforms. An important aspect is that the resulting waveforms can be represented by an amplitude and phase for each mode, with only a few hundred samples~\cite{Puerrer14}.

A complete representation of spin effects in fast IMR templates still remains a frontier of gravitational wave signal modelling. In presence of misaligned spins, the system will endure precession of the orbital plane as it evolves, leading to modulations of the signal as seen by the observer~\cite{ACST94}. Precessing EOB templates incorporate all six parameters of the two spins~\cite{Pan+13, BTB16} but still require a costly time-domain integration; while precessing phenomenological templates~\cite{Hannam+13} decrease the dimensionnality by using effective spin parameters representing only the dominant effects of the spins. To model the effect of precession on the emitted waveform, it has been proposed~\cite{BCV03b, BCPTV05, Schmidt+10, OShaughnessy+11, Boyle+11} to decompose precessing waveforms by performing a time-dependent rotation, following the precession of the orbital plane. The resulting waveform in the rotated frame can then be modelled by a non-precessing waveform, an approximation which is used both in the construction of the inspiral part of precessing EOB waveforms~\cite{Pan+13} and in the construction of precessing phenomenological waveforms~\cite{Hannam+13}. To follow this modelling approach and efficiently create Fourier-domain waveforms, one needs to understand how to translate the time-domain modulations created by the frame rotation into a Fourier-domain transfer function, which is our first motivation.

Prospective parameter recovery studies for LISA observations of comparable-mass binary black hole coalescences aim at characterizing the scientific performance of the future experiment. Although LISA~\cite{LISA17} is still at least a decade from launch, the ESA L3 mission is rapidly progressing into its early stages where mission design choices must be informed by science impact analysis. However, the recent progresses in waveform modelling, most notably the availability of fast and accurate IMR waveforms, have not yet made their way to the field of LISA data analysis studies. In the past, many such studies used post-Newtonian, inspiral-only models, and were limited to the cheaper Fisher-matrix formalism instead of a full Bayesian analysis (among numerous works, see e.g. \cite{Cutler97,LISAPE09}). Refs.~\cite{McWilliams+09,McWilliams+10,McWilliams+11} explored the effects of including the merger and ringdown, keeping the mass parameters fixed. However, the Fisher matrix analysis, although considerably cheaper to perform, is valid only in the limit of a high signal to noise ratio (SNR). Several implementations of Bayesian analyses were developed~\cite{CP06,Wickham+06,Babak+08,GP09,Feroz+09,Petiteau+10,PC13} (see also~\cite{MLDC4} and references therein), with Ref.~\cite{PC15} proposing a direct comparison of Fisher and Bayes analyses, but they were limited to inspiral-only waveforms. One exception was Ref.~\cite{Babak+08} where IMR signals were used in a Bayesian study, but there the masses of the black holes were kept fixed. In the recent Ref.~\cite{Klein+15}, the limitation to inspiral-only precessing post-Newtonian waveforms was compensated by an SNR reweighting procedure. Thus, due mainly to their high computational cost, full-fledged Bayesian parameter estimation simulations for IMR signals are still missing. \jgb{[Maybe there is too much detail in this paragraph?]}

@@

In contrast with current ground-based detectors, which are typically sensitive to chirping binaries for less than a second, LISA signals may accumulating over months or years.
The response of a LISA-type instrument is thus time-dependent~\cite{Cutler97}.
The motion and change of orientation of the detector constellation along its orbit lead to significant time variability in the form of a modulation and a varying delay. These effeccts then convey information about the localization of the gravitational-wave source in the sky. The complicated detector response increase the computational cost of parameter-estimation analyses. To leverage the performance of state-of-the-art Fourier-domain IMR templates~\cite{BTB16,Khan+15}, one needs to be able to process the signals through the time-dependent response of the dectector while staying in the Fourier domain.

@@

In previous works, the Stationary Phase Approximation (SPA) (see e.g.~\cite{Thorne300, CF94}) has been often used for this purpose. The SPA is a common approximation to compute the Fourier transform of non-precessing signals during the inspiralling phase. In the case of precessing binaries, applying it directly to the modulated signal is however prone to pathologies. In Ref.~\cite{KCY13,KCY14}, a formalism (called shifted uniform asymptotics or SUA) was introduced to go beyond the SPA and compute more accurately the modulation in the Fourier domain; however, this formalism still relies on the SPA for the underlying precessing-frame signal, and is as such limited to inspiralling signals. Another intuitive approach consists in using a time-to-frequency correspondence, such as the one given by the SPA, to directly carry over the modulation to the Fourier domain. This corresponds, in a sense to be detailed below, to the zeroth-order approximation of the SUA, and it is used implicitly\footnote{With the difference that the expressions used for the modulation are given in terms of frequency directly for the inspiral, and are then extended to also cover high frequencies in the merger-ringdown part of the signal, as we will discuss below.} for the simplified precession of the phenomenological waveforms of Ref.~\cite{Hannam+13}. In the case of the LISA response, the SPA provides a natural map from time-domain (for the orbit) to Fourier-domain (for the signal)~\cite{Cutler97}. Ref.~\cite{Klein+15} included the orbital motion of the detector in the SUA treatment. Consistently extending these previous approaches to the merger-ringdown part of the signals, including the delays in the LISA case, is part of the objectives of this paper.

The plan of the paper is as follows. In Sec.~\ref{sec:motivation}, we motivate the problem of Fourier-domain modulation and introduce the relevant timescales for both the response of LISA-type detectors and the modulation of precessing signals. In Sec.~\ref{sec:formalism}, we present our general formalism, give its leading order approximation as well as higher-order corrections, introduce new timescales based on the Fourier-domain signal, refine the previous results for both the quadratic-in-phase corrections and the treatment of the delays, and make the connection to previous approaches of the problem. We then apply our formalism to the response of the LISA detector in Sec.~\ref{sec:LISA}, and to the case of signals from precessing binaries in Sec~\ref{sec:precession}. We discuss and summarize our results in Sec.~\ref{sec:discussion}.

\SM{[talk about slowly chirping systems in the intro]}

@@

\section{Motivation}
\label{sec:motivation}

@@

\subsection{Basic considerations}
\label{subsec:basic}

We can jointly consider challenges in the contexts of both the LISA response and GW precession applications in terms of a more general problem which covers both domains. In each case must determine the Fourier transform of a signal $h(t)$ to which we apply a varying delay $d(t)$ before multiplication by a modulation function $F(t)$,
\be
\label{eq:delay-mod-defs}
	h_{d}(t) = h(t+d(t)) \,, \quad s(t) = F(t)h_{d}(t) \,,
\ee
and seek an efficient way to commpute the Fourier transform $\tilde{s}(f)$, expressed by means of a Fourier-domain transfer function $\calT$ such that
\be\label{eq:deftransfer}
	\tilde{s}(f) \equiv \calT(f) \tilde{h}(f) \,. 
\ee
As is well known, GW signals decomposed in spin-weighted spherical harmonic components are smoothly varying functions of frequency in amplitude/phase form. An important consequence is that these signal components can be accurately represented in terms of a relatively coarsely sampled frequency grid. This property will also extend to the transfer functions, allowing us to keep a compact representation of the full signals. (See Appendix~\ref{app:notation} for details of our notation and conventions for Fourier transforms and spherical harmonics.)

In the case of precessing binaries, the delays will be absent and the modulation functions will be the time-dependent Wigner coefficients applied to rotate the waveform from a precessing frame, where it verifies a restored mode hierarchy, to an inertial frame where the observations take place. In the case of a LISA-type detector, the signal $h(t)$ will simply be the waveform in a fixed heliocentric frame, and the delays will come from the motion of each detector in the wave front, while the modulation will represent the time-variation in the detector orientation.

@@

Together, these observations let us think that, for a given $f$, only a limited range of times should be relevant in $d(t)$ and $F(t)$ so that we may expect to find a treatment for the transfer function that would be local in time for the modulation and delay. Intuitively, for an inspiral signal and in the limit of an extremely slowly varying $d(t)$ and $F(t)$, we should be able to simply evaluate them at the time given by the time-to-frequency correspondence of the SPA. Furthermore, a natural quantitative criterion for the applicability of this idea is given by the comparison of the radiation-reaction timescale of the signal with the modulation timescale (possibly augmented by a dimensionless factor when delays are present). In other words, the change in frequency of the signal over a characeteristic time of the modulation should be small for the separation of timescales to work.

Our objective is to find an approximate treatment for $\calT(f)$ which allows us to exploit these properties without relying on unnecessary limiting assumptions for the signal (like the limitation of the SPA to inspiral signals), which is extensible to the high-accuracies which will be required by LISA and other future gravitational-wave instruments, while being computationally efficient and widely applicable to GW analysis. Before introducing developing our formalism, though, we review the salient features of our application problems.

@@

The response of a detector of the LISA type to an incident gravitational wave can be written in two different, equivalent forms, in terms of phase or frequency measurements. Here we will work with the second representation, which will prove more convenient for our purposes. Moreover, various notation and conventions have been used in the literature to label the spacecrafts and describe their orbits. We refer the reader to~\cite{Vallisneri04} for a comparative account on these various conventions. In this work, we will keep close to the conventions of~\cite{Vallisneri04}, which were also used in the Mock LISA Data Challenges (MLDC)~\cite{MLDC4}.

We use a coordinate system centered on the solar system barycenter (SSB), and represent the center of the constellation by the vector $p_{0}$. We introduce the notation $\hatk$ for the propagation vector of the gravitational wave, which we denote by $h_{ij}^{\rm TT}(t)$ in transverse-traceless matrix form, as measured at the SSB (thus, at position $p$, $h_{ij}^{\rm TT}(t, p) = h_{ij}^{\rm TT}(t - \hatk \cdot p)$). We denote by $p_{A}$ ($A=1,2,3$) the position of the individual spacecrafts and $n_{l}$ the unit vectors of the three links, with the convention that $n_{3}$ points from 1 to 2.

Written in terms of the fractional laser frequency shifts between two spacecrafts, the elementary response of the detector reads~\cite{EW75, RCP04, Vallisneri04}
\begin{align}\label{eq:yslr}
	y_{slr} &\equiv \frac{\nu_{r} - \nu_{s}}{\nu} \nn\\
	&= \frac{1}{2} \frac{n_{l}^{i}n_{l}^{j}}{1 - \hatk\cdot n_{l}} \left[ h_{ij}^{\rm TT}(t - L - \hatk\cdot p_{s}) - h_{ij}^{\rm TT}(t - \hatk\cdot p_{r}) \right] \,.
\end{align}
We will use a rigid instantaneous model, approximating the geometry of the constellation by a moving equilateral triangle (neglecting the flexing of the arms induced by corrections in the orbits), and evaluating all geometric factors in~\eqref{eq:yslr} at a single time $t$ (neglecting pointing-ahead corrections). The additional delay $L$ (we use $c=1$) in the first term represents the light propagation time along the arm between spacecrafts $s$ and $l$.

The first- and second-generation TDI observables are then built as combinations of these basic building blocks, evaluated at delayed times. Since in our rigid approximation these delays will take a simple form in Fourier domain, we will focus in this paper on the $y_{slr}$ observables. We will give in Sec.~\ref{subsec:modelLISA} more details on the response of LISA-type detectors, on the TDI observables, and on the approximations that enter the derivation of the basic response~\eqref{eq:yslr} above.

@@

The functions $F(t)$ and $d(t)$ vary on a timescale of one year, with a frequency $f_{0} = 1/\mathrm{yr} \simeq 3.169\times10^{-8} \mathrm{Hz}$ (we will also use $\Omega_{0} = 2\pi f_{0}$). When neglecting small non-periodic orbital perturbations (such as those caused by the influence of the Earth and of the other planets on the orbits), both $F(t)$ and $d(t)$ are periodic. It will also be useful to separate the delays in two types of terms: the first, $d_{0} = -\hatk\cdot p_{0}$, relates the waveform at the SSB to the waveform at the center of the constellation, whereas the second, $d_{c}$, represents the various delays between the spacecrafts of the constellation. For an orbit radius $R=1\,\mathrm{au}$ and an armlength $L=2.5\times10^{6}\mathrm{km}$~\cite{LISA17}, $d_{0} \sim R/c \simeq 500\mathrm{s}$ and $d_{c} \sim L/c \simeq 8\mathrm{s}$.

@@

For spinning compact objects, the presence of the spins leads to the precession of the orbit~\cite{Apostolatos+94, Kidder95}, which can have large effects on the waveform. In particular, it breaks the planar symmetry of the gravitational wave emission, and leads to modulations that are especially important for systems that are observed edge-on.

It was suggested by a number of authors~\cite{BCV03b, BCPTV05, Schmidt+10, OShaughnessy+11, Boyle+11} that the effect of the precession can be modeled, to a good approximation, by a time-dependent rotation of a non-precessing waveform. This allows for a modelling approach where one models separately a precessing frame following the evolution of the plane of the orbit, and approximates the waveform in this precessing frame by using a non-precessing model. Different prescriptions have been proposed for the construction of a precessing frame from the waveform itself~\cite{Schmidt+10, OShaughnessy+11, Boyle+11}.

@@

The frequency-shift response for a single link~\eqref{eq:yslr} was derived in Ref.~\cite{EW75} (see also~\cite{RCP04, Finn08, Cornish09} for more modern derivations). Several assumptions enter the result as written in~\eqref{eq:yslr}: (i) effects of the order $v/c$ are neglected, including for instance the special relativistic Doppler effect created by the relative speeds of the spacecrafts on their orbits (ii) the propagation is assumed to take place in a flat spacetime, perturbed only by the gravitational wave; thus the gravitational redshift as well as the deflection of light created by the gravitational potential of the Sun is ignored (iii) all geometric factors are evaluated at a single time, whereas one should consider the beam as propagating from the position of the first spacecraft at the time of emission to the position of the second scapecraft at the time of reception, leading to a pointing-ahead effect.

>>>>

\jgb{With the unprecedented recent gravitational-wave detections of coalescencing binary black holes and binary neutron stars, announced by the LIGO-Virgo collaboration~\cite{LIGO-theevent-2016,LIGO-christmasevent-2016,LIGO-O1BBH-2016}[update refs], gravitational-wave astronomy has entered its observational era. As LIGO prepares for even more sensitive observation runs, and with the recent expansion of the ground-based detectors network with Virgo~\cite{Virgo} ( and eventually also KAGRA~\cite{KAGRA} and LIGO-India~\cite{INDIGO}), observations of such compact object coalescences are expected at ever-increasing rate, with binary black holes likely to be the dominant sources~\cite{} \SM{[C]}.}

  \jgb{Moreover, the European Space Agency has recently selected the Laser Interferometer Space Antenna (LISA)\cite{LISA17} to realize the ``Gravitational Universe'' science theme\cite{GravitationalUniverse2013} as the 3rd large space mission of its Cosmic Vision program, with a tentative launch around 2034. LISA will be able to detect and characterize, among several important gravitational wave source targets, comparable-mass binary black hole coalescences from cosmological distances over a wide range of masses. These will range from high-redshift observations of supermassive black hole binaries with $M\sim 10^{7} \Msol$ down to the observation of LIGO-type sources with $M\sim 10^{2} \Msol$~\cite{Sesana16}.}

  Data analysis for gravitational-wave observations of compact binary coalescences require accurate models (or templates) for the signals, both for ensuring efficient detections of signals that may be buried in instrumental noise, and to extract the physical parameters of the source in a subsequent analysis. Bayesian analysis for parameter estimation of gravitational-wave signals, as was performed for the LIGO detections~\cite{LIGO-theeventPE-2016,LIGO-O1BBH-2016}, may require millions of evaluations of the likelihood function to sample the posterior probability distribution. \SM{[objective: accuracy, not necessary necessary for LIGO now but in the future yes]} The greater sensitivity of LISA and other future instruments will require further increases in the accuracy and computational efficiency of signal templates.

  %This motivates an important and ongoing effort to improve gravitational wave templates. Two important focus areas for template improvements are need to cover all phases of the evolution of the system, including the inspiral, merger and ringdown (IMR), which became possible thanks to the breakthrough of numerical relativity~\cite{Pretorius05,Baker+06,Campanelli+06}, and the inclusion of the effects of the spins of the compact objects, that can lead to a precession of the orbital plane~\cite{ACST94,K95}. Besides accuracy, the computational efficiency of template generation is crucial.

\jgb{The GW community is making progress assembling higher-fidelity tools for these coming challenges.
State-of-the-art IMR templates combine information from the perturbative results of post-Newtonian theory covering the inspiral (see e.g.~\cite{BlanchetLiving}) and from numerical relativity simulations covering the end of the inspiral and the merger-ringdown phase (see e.g.~\cite{Pfeiffer12}).  Approaches to template construction include phenomenological templates postulating an analytic ansatz for the Fourier-domain amplitude and phase~\cite{Husa+15, Khan+15,Hannam+13}, and the more physically motived Effective-One-Body (EOB) approach~\cite{BD99,Taracchini+13, Pan+13, Bohe+16}. Where needed, Reduced Order Models (ROM), also called surrogate models, have been developed to considerably speed up waveform generation, without losing accuracy~\cite{Field+13, Puerrer14, Bohe+16}. These tools provide efficient non-precessing IMR Fourier-domain waveforms. Importantly for this work, the resulting waveforms can be represented by an amplitude and phase for each mode, with only a few hundred samples~\cite{Puerrer14}.[JB:I cut this down significantly.]} Our challenge will be to accurately complete the representation of instrumental signals including precession effects and the LISA instrumental response and precession effects without sacrificing the computational efficiency enabled by these fast waveform templates.

A complete representation of spin effects in fast IMR templates still remains a frontier of gravitational wave signal modelling. In presence of misaligned spins, the system will endure precession of the orbital plane as it evolves, leading to modulations of the signal as seen by the observer~\cite{ACST94}.
%Precessing EOB templates incorporate all six parameters of the two spins~\cite{Pan+13, BTB16} but still require a costly time-domain integration; while precessing phenomenological templates~\cite{Hannam+13} decrease the dimensionnality by using effective spin parameters representing only the dominant effects of the spins.
\jgb{[JB:I cut this down too.]}

A promising approach to modeling the effect of precession on the emitted waveform, it has been proposed~\cite{BCV03b, BCPTV05, Schmidt+10, OShaughnessy+11, Boyle+11} to decompose precessing waveforms by performing a time-dependent rotation, following the precession of the orbital plane. The resulting waveform in the rotated frame can then be modelled by a non-precessing waveform, an approximation which is used both in the construction of the inspiral part of precessing EOB waveforms~\cite{Pan+13} and in the construction of precessing phenomenological waveforms~\cite{Hannam+13}. To follow this modelling approach and efficiently create Fourier-domain waveforms, one needs to understand how to translate the time-domain modulations created by the frame rotation into a Fourier-domain transfer function, which is our first motivation.

Beyond a fast representation of the incident waveform, observational analyses also require transforming the signals through some instrumental response. For short duration mergers such as LIGO  has detected, this can be treated by a simple multiplier.  For future instruments though, the instrumental response will be more complicated.

%Prospective parameter recovery studies for LISA observations of comparable-mass binary black hole coalescences aim at characterizing the scientific performance of the future experiment. Although LISA~\cite{LISA17} is still at least a decade from launch, the ESA L3 mission is rapidly progressing into its early stages where mission design choices must be informed by science impact analysis. However, the recent progresses in waveform modelling, most notably the availability of fast and accurate IMR waveforms, have not yet made their way to the field of LISA data analysis studies. In the past, many such studies used post-Newtonian, inspiral-only models, and were limited to the cheaper Fisher-matrix formalism instead of a full Bayesian analysis (among numerous works, see e.g. \cite{Cutler97,LISAPE09}). Refs.~\cite{McWilliams+09,McWilliams+10,McWilliams+11} explored the effects of including the merger and ringdown, keeping the mass parameters fixed. However, the Fisher matrix analysis, although considerably cheaper to perform, is valid only in the limit of a high signal to noise ratio (SNR). Several implementations of Bayesian analyses were developed~\cite{CP06,Wickham+06,Babak+08,GP09,Feroz+09,Petiteau+10,PC13} (see also~\cite{MLDC4} and references therein), with Ref.~\cite{PC15} proposing a direct comparison of Fisher and Bayes analyses, but they were limited to inspiral-only waveforms. One exception was Ref.~\cite{Babak+08} where IMR signals were used in a Bayesian study, but there the masses of the black holes were kept fixed. In the recent Ref.~\cite{Klein+15}, the limitation to inspiral-only precessing post-Newtonian waveforms was compensated by an SNR reweighting procedure. Thus, due mainly to their high computational cost, full-fledged Bayesian parameter estimation simulations for IMR signals are still missing. \jgb{[Maybe there is too much detail in this paragraph?]}

@@

\jgb{Whereas LIGO and Virgo are typically sensitive to chirping binaries for a minute or less, LISA signals may accumulating over months or years.
The response of a LISA-type instrument is thus time-dependent~\cite{Cutler97}.
The motion and change of orientation of the detector constellation along its orbit lead to significant time variability in the form of a modulation and a varying delay. These effects then convey information about the localization of the gravitational-wave source in the sky. Direct implementation of the detector response is straightforward\cite{SynthLISA,LISACode,LISASim} but at prohibitive computational cost for parameter-estimation analyses. To leverage the performance of state-of-the-art Fourier-domain IMR templates~\cite{BTB16,Khan+15}, we must efficiently process the signals through the time-dependent response of the dectector while staying in the Fourier domain.}

@@

In previous works focused on gravitational-wave inspirals, the Stationary Phase Approximation (SPA) (see e.g.~\cite{Thorne300, CF94}) has been often used for this purpose. While the SPA is a common approximation to compute the Fourier transform of non-precessing signals during the inspiraling phase, it is not applicable for IMR waveforms.

\jgb{In the case of precessing binaries, applying the SPA directly to the modulated signal is also prone to pathologies. In Ref.~\cite{KCY13,KCY14}, a formalism (called shifted uniform asymptotics or SUA) was introduced to go beyond the SPA and compute more accurately the modulation in the Fourier domain; however, this formalism still relies on the SPA for the underlying precessing-frame signal, and is as such limited to inspiraling signals. \textbf{[JB: Still uncertain about the rest of this paragraph.]} Most directly, a time-to-frequency correspondence, such as the one given by the SPA, can be exploited to directly carry over the precision modulation to the Fourier domain.} This corresponds, in a sense to be detailed below, to the zeroth-order approximation of the SUA, and it is used implicitly\footnote{With the difference that the expressions used for the modulation are given in terms of frequency directly for the inspiral, and are then extended to also cover high frequencies in the merger-ringdown part of the signal, as we will discuss below.} for the simplified precession of the phenomenological waveforms of Ref.~\cite{Hannam+13}.

In the case of the LISA response, the SPA provides a natural map from time-domain (for the orbit) to Fourier-domain (for the signal)~\cite{Cutler97}. Ref.~\cite{Klein+15} included the orbital motion of the detector in the SUA treatment. Consistently extending these previous approaches to the merger-ringdown part of the signals, including the delays in the LISA case, is part of the objectives of this paper.

We seek to overcome two limitations in such previous approaches.  The first is that  SPA-based methods are not applicable to IMR waveforms. Second, there is often no clear way to improve the accuracy of these methods beyond the intuitive leading order treatment in order to meet the high-accuracy needs of future detectors. Our approach exploits separation of time-scales approximations, based on a general treatment directly in the Fourier domain of slowly varying delays and amplitude modulations for chirping waveforms with slowly varying amplitude and phase.

The plan of the paper is as follows. In Sec.~\ref{sec:motivation}, we provide a general presentation of problem of Fourier-domain modulation and introduce the relevant timescales for both the response of LISA-type detectors and the modulation of precessing signals. In Sec.~\ref{sec:formalism}, we present our general formalism, give its leading order approximation as well as higher-order corrections, introduce new timescales based on the Fourier-domain signal, refine the previous results for both the quadratic-in-phase corrections and the treatment of the delays, and make the connection to previous approaches of the problem. We then apply our formalism to the response of the LISA detector in Sec.~\ref{sec:LISA}, and to the case of signals from precessing binaries in Sec~\ref{sec:precession}. We discuss and summarize our results in Sec.~\ref{sec:discussion}.

@@

\section{GW signals in the frequency domain}
\label{sec:motivation}

@@

%\subsection{Basic considerations}
%\label{subsec:basic}

\jgb{To motivate our signal treatment we consider challenges of both the LISA response and GW precession applications in terms of a more general problem which covers both domains. In each case must determine the Fourier transform a signal $\tilde s(f)$ which results from processing some incident waveform $h(t)$ through some combination of modulations and delays.}

In the general problem, we apply a time-varying delay $d(t)$ to the waveform followed by a multiplicative modulation function $F(t)$,
\be
\label{eq:delay-mod-defs}
	h_{d}(t) = h(t+d(t)) \,, \quad s(t) = F(t)h_{d}(t) \,.
\ee
We then seek an \textit{efficient} way to compute the Fourier transform $\tilde{s}(f)$, expressed by means of a Fourier-domain transfer function $\calT$ such that
\be\label{eq:deftransfer}
	\tilde{s}(f) \equiv \calT(f) \tilde{h}(f) \,. 
\ee

As is well known, GW signals decomposed in spin-weighted spherical harmonic components are smoothly varying functions of frequency in amplitude/phase form\cite{smoothAmpPhaseRefs}. An important consequence is that these signal components can be accurately represented in terms of a relatively coarsely sampled frequency grid. This property will also extend to the transfer functions, allowing us to keep a compact representation of the full signals. (See Appendix~\ref{app:notation} for details of our notation and conventions for Fourier transforms and spherical harmonics.)

In the case of precessing binaries, the delays will be absent and the modulation functions will be the time-dependent Wigner coefficients applied to rotate the waveform from an \jgb{precessing orbital frame, in which waveform modes exhibit smooth amplitude and phase variation}, to an inertial frame where the observations take place. In the case of a LISA-type detector, the signal $h(t)$ will simply be the waveform in a fixed heliocentric frame, and the delays will come from the motion of each detector against the wave front, while the modulation will represent the time-variation in the detector orientation.

@@

Together, these observations lead us to expect that, for a given $f$, only a limited range of times should be relevant in $d(t)$ and $F(t)$ so that we may expect to find a treatment for the transfer function that would be local in time for the modulation and delay. This general idea has already been applied for inspiral signals in the limit of an extremely slowly varying $d(t)$ and $F(t)$, where intuitively we should be able to simply evaluate them at the time given by the time-to-frequency correspondence of the SPA.

Furthermore, a natural quantitative criterion for the applicability of this idea is given by the comparison of the radiation-reaction timescale of the signal with the modulation timescale. In other words, the change in frequency of the signal over a characeteristic time of the modulation should be small for the separation of timescales to work.

Our objective is to find an approximate treatment for $\calT(f)$ which allows us to exploit these properties without relying on unnecessary limiting assumptions for the signal (like the limitation of the SPA to inspiral signals), which is extensible to the high-accuracies which will be required by LISA and other future gravitational-wave instruments, while being computationally efficient and widely applicable to GW analysis. In preparation for developing our formalism we first review the salient features of our application problems.

@@

The response of a detector of the LISA type to an incident gravitational wave can be written in two different, equivalent forms, in terms of phase or frequency measurements. Here we will work with the second representation, which will prove more convenient for our purposes. Moreover, various notation and conventions have been used in the literature to label the spacecraft and describe their orbits. We refer the reader to~\cite{Vallisneri04} for a comparative account on these various conventions. In this work, we will keep close to the conventions of~\cite{Vallisneri04}, which were also used in the Mock LISA Data Challenges (MLDC)~\cite{MLDC4}.

We use a coordinate system centered on the solar system barycenter (SSB), and represent the center of the constellation by the vector $p_{0}$. We introduce the notation $\hatk$ for the propagation vector of the gravitational wave, which we denote by $h_{ij}^{\rm TT}(t)$ in transverse-traceless matrix form, as measured at the SSB (thus, at position $p$, $h_{ij}^{\rm TT}(t, p) = h_{ij}^{\rm TT}(t - \hatk \cdot p)$). We denote by $p_{A}$ ($A=1,2,3$) the position of the individual spacecraft and $n_{l}$ the unit vectors of the three links, with the convention that $n_{3}$ points from 1 to 2.

Written in terms of the fractional laser frequency shifts between two spacecraft, the elementary response of the detector reads~\cite{EW75, RCP04, Vallisneri04}
\begin{align}\label{eq:yslr}
	y_{slr} &\equiv \frac{\nu_{r} - \nu_{s}}{\nu} \nn\\
	&= \frac{1}{2} \frac{n_{l}^{i}n_{l}^{j}}{1 - \hatk\cdot n_{l}} \left[ h_{ij}^{\rm TT}(t - L - \hatk\cdot p_{s}) - h_{ij}^{\rm TT}(t - \hatk\cdot p_{r}) \right] \,.
\end{align}
We will use a rigid instantaneous model, approximating the geometry of the constellation by a moving equilateral triangle (neglecting the flexing of the arms induced by corrections in the orbits), and evaluating all geometric factors in~\eqref{eq:yslr} at a single time $t$ (neglecting point-ahead corrections). The additional delay $L$ (we use $c=1$) in the first term represents the light propagation time along the arm between spacecraft $s$ and $l$.

The first- and second-generation TDI observables are then built as combinations of these basic building blocks, evaluated at delayed times. Since in our rigid approximation these delays will take a simple form in Fourier domain, we will focus in this paper on the $y_{slr}$ observables. Sec.~\ref{subsec:modelLISA} provides more details on the response of LISA-type detectors, on the TDI observables, and on the approximations that enter the derivation of the basic response~\eqref{eq:yslr} above.

@@

For the LISA response, the functions $F(t)$ and $d(t)$ vary on a timescale of one year, with a frequency $f_{0} = 1/\mathrm{yr} \simeq 3.169\times10^{-8} \mathrm{Hz}$ (we will also use $\Omega_{0} = 2\pi f_{0}$). When neglecting small non-periodic orbital perturbations (such as those caused by the influence of the Earth and of the other planets on the orbits), both $F(t)$ and $d(t)$ are periodic. It will also be useful to separate the delays in two types of terms: the first, $d_{0} = -\hatk\cdot p_{0}$, relates the waveform at the SSB to the waveform at the center of the constellation, whereas the second, $d_{c}$, represents the various delays between the spacecraft of the constellation. \jgb{Concretely, we assume the basic reference design parameters of the LISA mission\cite{LISA17} recently selected by the European Space Agency(ESA), with orbit radius $R=1\,\mathrm{au}$ and armlength $L=2.5\times10^{6}\mathrm{km}$, and with $d_{0} \sim R/c \simeq 500\mathrm{s}$ and $d_{c} \sim L/c \simeq 8\mathrm{s}$.}

@@

For spinning compact objects, \jgb{angular momentum interactions typically} lead to the precession of the orbit~\cite{Apostolatos+94, Kidder95}, which can have large effects on the waveform. In particular, it breaks the planar symmetry of the gravitational wave emission, causing modulations that are especially important for systems that are observed edge-on.

A number of authors~\cite{BCV03b, BCPTV05, Schmidt+10, OShaughnessy+11, Boyle+11} have suggested that the effect of the precession can be modeled, to a good approximation, by a time-dependent rotation of a effectively non-precessing waveform. This allows for a modelling approach where one separately models a precessing frame following the evolution of the plane of the orbit, and approximates the waveform in this precessing frame by using a effective non-precessing model. Different prescriptions have been proposed for the construction of a precessing frame from the waveform itself~\cite{Schmidt+10, OShaughnessy+11, Boyle+11}.

@@

The frequency-shift response for a single link~\eqref{eq:yslr} was derived in Ref.~\cite{EW75} (see also~\cite{RCP04, Finn08, Cornish09} for more modern derivations). Several assumptions enter the result as written in~\eqref{eq:yslr}: (i) effects of the order $v/c$ are neglected, including for instance the special relativistic Doppler effect created by the relative speeds of the spacecraft on their orbits (ii) the propagation is assumed to take place in a flat spacetime, perturbed only by the gravitational wave; thus the gravitational redshift as well as the deflection of light created by the gravitational potential of the Sun is ignored (iii) all geometric factors are evaluated at a single time, whereas one should consider the beam as propagating from the position of the first spacecraft at the time of emission to the position of the second scapecraft at the time of reception, leading to a pointing-ahead effect.
